How to select predictive models for decision-making or causal inference?

### 2024-10-06 response_to_reviewer.md

Let me begin with positive comments.  I think the paper is very important because it describes a sophisticated methodology for experimental examination of causal models.
Despite some attempts this is still lacking in literature.  I especially like the simulation scheme which uses different features for data generation and learning models.  Moreover thorough experimental evaluations of causal models are not very frequent making the paper important.  Code is available on Github which is a big plus.

I am however not sure about the overall conclusions, which are that R-risk is always the best.  Performance of risk functions may depend on several factors and the overall best performance might be a result of bias in available data.  The only factor taken into account by the authors is overlap.  However, only one type of lack of overlap is tested in experiments, based on shifted Gaussians.  This will likely lead to violation of the positivity assumption.  How about different scenarios, e.g.: perfect overlap but different treatment/control distributions?  Confounding need not necessarily lead to lack of overlap.  Another possible factor is the ratio of causal effect to background responses.  It is known that it may affect relative performance of HTE estimators.

Overall: I am no fully convinced that R-risk is always the best.  The answer could be dependent on other factors and I hope the authors would explore more scenarios.

> We thank the reviewer for this remark and the suggestion to explore more diverse simulation scenarii. Exploring the sensitivity of our result on the R-risk appears to us particularly necessary. Thus, we ran a dedicated experiment focusing on the ratio of causal effect to background response. We measure the ratio by the absolute mean difference between the causal effect and the baseline response, measured empirically on each dataset instance $\Delta_{\mu} = \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i)} \big|$. We vary a dedicated parameter ($\omega$ in Algorithm 2) in the simulation to force this ratio to cover a wide range of values. In 90 dataset instance, it ranges several orders of magnitude from 0.005 to 9675. Figure 7 gives the main results of this experiment. We see that R-risk is still the best risk on average, but when the effect ratio is high (tertile between 9 and 9675), the reweighted mu-risk and the the tau-risk_IPW are better than the R-risk. However, for low value of the ratio,  the R-risk largely outperforms all other metrics. We added these new results to the main paper. We also nuanced our conclusion on the R-risk, underlying that it is better on average but that in some situations, other risks might be more adapted. A typical such situation would be when the applied scientist is interested in estimating an effect with a big ratio and want to focus on the part of the population where there is low overlap.

@@

> The second paper is from the uplift regression literature. It focus on full randomization problems and a linear specification with gaussian predictors. It shows theoretically that a linear T-learner (called double model approach in their approach) has smaller asymptotical variance than a treatment ratio reweighted outcome S learner. However, they note that in non normal cases the  T-learner can be ill-defined or with high variance. They propose a corrected reweighted S-learner approach to reduce this  variance. They show empirically that it behaves better than the non corrected version on a simulation and two real world problems. 
> I am not sure to see the link with the tau-risk. The paper does not compare risks but corrected or uncorrected vs T-learners. In addition, they only consider full randomization and linear problems. In our case, we consider a more general setting with a variety of data generation procedures and we focus on non-randomized situations where the distribution of the treatment is conditioned by the predictors.

Another problem is that the paper is hard to follow. The authors rapidly switch between topics: risk, selection, nuisance parameter estimation so the reader quickly gets lost.  The sections in supplementary materials contain important information but not all are referenced in main text.  The paper gives an impression of several experiments put together in some order without guiding the reader through the maze.  I think organization should be improved before publication.

> We tried to better guide the reader through the method section and to clarify the motivation behind the different experiments. 
> We checked the supplementary materials but we did not find any appendix that was not referenced at least once in the main text (several are referred in figure captions). Did you have a specific appendix in mind that would require a reference in the main text?

### 2024-10-06 how_to_select_latex.tex
effect ratio: include result of expe
- diff with original publication
- answer to reviewer: complete with new results

\paragraph{Robinson decomposition}
G-computation is one choice of decomposition of the CATE estimation into a sequence of regressions, where the only statistical estimate is the response function $\mu_{a}(x)$ \cite[chapter 15]{chernozhukov2024applied}. Other choices of decomposition exist, such as the R-decomposition \cite{robinson_rootnconsistent_1988}. It introduces two new statistical estimates, the conditional mean outcome
and the probability to be treated (known as propensity score \cite{rosenbaum_central_1983}):
\begin{flalign}
    \text{Conditional mean outcome} &                    & m(x) \myeq \; & \mathbb E_{Y \sim
    \mathcal{D}} [Y|X=x]            &                    &
    \label{def:m}
    \\
    \text{Propensity score}         &                    &
    e(x) \myeq \;                   & \mathbb P[A=1|X=x]
    \label{def:propensity_score}
\end{flalign}
the outcome can be written
\begin{flalign}\label{eq:r_decomposition}
    \text{R-decomposition}            &  & y(a) = m(x) + \big( a - e(x) \big)
    \tau(x) + \varepsilon(x; a)\notag &  &
    \\\ &&\text{with}\quad \mathbb E[\varepsilon(X; A)|X, A] = 0&&
\end{flalign}
$m$ and $e$ are often called
\emph{nuisances} \cite{chernozhukov_double_2018}. They are unknown and are estimated in practice thanks to statistical models (eg. linear models or forest based estimators).

@@

\paragraph{R-risk is robust to a wide range of effect ratio}

We focused on the

\begin{figure}[!tb]
    % XXX: for submission
    \centering\begin{minipage}{\linewidth}
        \includegraphics[width=\linewidth]{_6_effect_ratio_by_bin_comparaison_kendall_by_Dataset.pdf}
    \end{minipage}

    \caption{\textbf{R-risk is robust to a wide range of effect ratio}: Kendall’s $\tau$ agreement with $\tau$-risk. Strong, medium and Weak Causal effect ratio are the tertiles of the absolute ratio causal effect to baseline response, ranging from 1 to 300. Appendix \ref{apd:effect_ratio} details this simulation.}\label{fig:effect_ratio_influence}
\end{figure}

@@

\paragraph{Nuisance models: more gain than pain}
%
Predictive models are increasingly used to reason about treatment effects, for
instance in precision medicine to drive individualized decision. Our results
highlight that they should be selected, validated, and tuned using different
procedures and error measures than those classically used to assess prediction. Rather, selecting the best outcome
model according to the $R\text{-risk}$ (eq.\, Definition \ref{def:r_risk}) leads
to more valid causal estimates. Estimating the $R\text{-risk}$ requires a more
complex procedure than standard cross-validation used \eg in machine
learning: it involves fitting nuisance models necessary for model
evaluation.
Our results show that these can be learned on the same set of data as the
outcome models evaluated. The nuisance models must be well estimated (Figure
\ref{fig:all_datasets_nuisances_comparison}). Our results show that using for
nuisance models a flexible stacking-based family of estimator suffices for good
model selection. To select propensity score models, we used the Brier score,
minimized by the true individual probability. An easy mistake is to
use calibration errors popular in machine learning
\cite{platt_probabilistic_1999,zadrozny_obtaining_2001,niculescu-mizil_predicting_2005,minderer_revisiting_2021}
as these select not for the individual posterior probability but for an
aggregate error rate \cite{perez2022beyond}.

@@

Increasingly complex prediction models integrating richer medical data
have flourished because their predictions can be easily
demonstrated and validated on left-out data. But using them to underpin a
decision on whether to treat or not requires more careful validation,
using a metric accounting for the
putative intervention, the $R\text{-risk}$. On average, the $R\text{-risk}$
brings a sizeable benefit to select the most adequate model, even when
model development is based on treated and
untreated population with little differences, as in RCTs. To facilitate better model selection, we provide Python
code\footnote{\url{https://github.com/soda-inria/causal_model_selection}}.
% Using the $R\text{-risk}$ does make evaluation more
% complicated not only because the procedure is more involved, but also
% because each intervention requires a dedicated evaluation. However, such
% off-policy evaluation remains much less costly than the recommended good
% practice of impact evaluation testing the ability of a prediction model
% to actually guide patient health \cite{hendriksen2013diagnostic}. Also,
This model-selection procedure puts no constraints on the models used to
build predictive models: it opens the door to evaluating a wide range of
models, from gradient boosting to convolutional neutral, or language
models.

\paragraph{Research question needing more work}

On average, the R-risk outperforms other risks, but it is not optimal for every
situations. If one possess some prior knowledge on the data generation process,
another risk might be more adapted. This is typically the case for big causal
effect ratio to baseline and low overlap where the $\tau\text{-risk}_{IPW}$
should be better.

@@

TODO

\bibliography{biblio}
%\printbibliography

>>>>

\paragraph{Robinson decomposition}
G-computation is a choice of decomposition of the CATE estimation into a
sequence of regressions, where the only statistical estimate is the response
function $\mu_{a}(x)$ \cite[chapter 15]{chernozhukov2024applied}. Other choices
of decomposition exist, such as the R-decomposition
\cite{robinson_rootnconsistent_1988}. The latter introduces two new statistical
estimates, the conditional mean outcome and the probability of being treated (known
as propensity score \cite{rosenbaum_central_1983}):
\begin{flalign}
    \text{Conditional mean outcome} &                    & m(x) \myeq \; & \mathbb E_{Y \sim
    \mathcal{D}} [Y|X=x]            &                    &
    \label{def:m}
    \\
    \text{Propensity score}         &                    &
    e(x) \myeq \;                   & \mathbb P[A=1|X=x]
    \label{def:propensity_score}
\end{flalign}
the outcome can be written
\begin{flalign}\label{eq:r_decomposition}
    \text{R-decomposition}            &  & y(a) = m(x) + \big( a - e(x) \big)
    \tau(x) + \varepsilon(x; a)\notag &  &
    \\\ &&\text{with}\quad \mathbb E[\varepsilon(X; A)|X, A] = 0&&
\end{flalign}
$m$ and $e$ are often called
\emph{nuisances} \cite{chernozhukov_double_2018}. They are unknown and are estimated in practice thanks to statistical models (eg. linear models or forest based estimators).

@@

\paragraph{R-risk is robust to a wide range of effect ratio values}

Beyond overlap, we study for caussim simulations, the effect on model selection
of different causal effect ratio to baseline. We vary the empirical mean
absolute difference between the causal effect and the baseline, $\Delta_{\mu} =
    \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i)}
    \big|$, covering a ratio range from 0.05 to 1000 ($\text{median}=1.9$, full
distribution in Appendix \ref{apd:effect_ratio}). Figure
\ref{fig:effect_ratio_influence} shows that for high values of the ratio,
$R-\text{risk}$ is outperformed by the $\mu\text{-risk}_{IPW}$ and the
$\tau\text{-risk}_{IPW}$. However, on average, the $R-\text{risk}$ is still the
better risk.

\begin{figure}[!tb]
    \centering
    \begin{minipage}{\linewidth}
        \includegraphics[width=0.9\linewidth]{_6_effect_ratio_by_bin_comparaison_kendall_by_Dataset.pdf}
    \end{minipage}

    \caption{\textbf{R-risk is robust to a wide range of effect ratio}:
    Kendall's $\tau$ agreement with $\tau$-risk. Strong, medium and Weak Causal
    effect ratio are the tertiles of the absolute ratio causal effect to
    baseline response, $\Delta_{\mu}$: Low $[0.005;0.4[$, Medium $[0.44; 9[$,
    High $[9;9675[$. Appendix \ref{apd:effect_ratio} details this
    simulation.}\label{fig:effect_ratio_influence}
\end{figure}

@@

\paragraph{Nuisance models: more gain than pain}
%
Predictive models are increasingly used to reason about treatment effects, for
instance in precision medicine to drive individualized decision. Our results
highlight that they should be selected, validated, and tuned using different
procedures and error measures than those classically used to assess prediction. Rather, selecting the best outcome
model according to the $R\text{-risk}$ (eq.\, Definition \ref{def:r_risk}) leads
to more valid causal estimates on average. Estimating the $R\text{-risk}$ requires a more
complex procedure than standard cross-validation used \eg in machine
learning: it involves fitting nuisance models necessary for model
evaluation.
Our results show that these can be learned on the same set of data as the
outcome models evaluated. The nuisance models must be well estimated (Figure
\ref{fig:all_datasets_nuisances_comparison}). Our results show that using for
nuisance models a flexible stacking-based family of estimator suffices for good
model selection. To select propensity score models, we used the Brier score,
minimized by the true individual probability. An easy mistake is to
use calibration errors popular in machine learning
\cite{platt_probabilistic_1999,zadrozny_obtaining_2001,niculescu-mizil_predicting_2005,minderer_revisiting_2021}
as these select not for the individual posterior probability but for an
aggregate error rate \cite{perez2022beyond}.

@@

Increasingly complex prediction models integrating richer medical data
have flourished because their predictions can be easily
demonstrated and validated on left-out data. But using them to underpin a
decision on whether to treat or not requires more careful validation,
using a metric accounting for the
putative intervention, the $R\text{-risk}$. On average, the $R\text{-risk}$
brings a sizeable benefit to select the most adequate model, even when
model development is based on treated and
untreated population with little differences, as in RCTs. To facilitate better model selection, we provide Python
code\footnote{\url{https://github.com/soda-inria/causal_model_selection}}.
% Using the $R\text{-risk}$ does make evaluation more
% complicated not only because the procedure is more involved, but also
% because each intervention requires a dedicated evaluation. However, such
% off-policy evaluation remains much less costly than the recommended good
% practice of impact evaluation testing the ability of a prediction model
% to actually guide patient health \cite{hendriksen2013diagnostic}. Also,
This model-selection procedure puts no constraints on the models used to
build predictive models: it opens the door to evaluating a wide range of
models, from gradient boosting to convolutional neural network, or language
models.

\paragraph{Limitations}

On average, the R-risk outperforms other risks, but it is not optimal for every
situations. If one possesses some prior knowledge on the data generation process,
another risk might be more adapted. This is typically the case for big causal
effect ratio to baseline and low overlap where the $\tau\text{-risk}_{IPW}$
should be better.

@@

Going beyond overlap, we study the effect of another important parameter of the
data generation process: the ratio of causal effect to the baseline response.

For each dataset of our simulation (caussim), we measure this ratio as the absolute mean
difference between the causal effect and the baseline response, measured
empirically on each dataset instance $\Delta_{\mu} = \frac{1}{N} \sum_{i=1}^N
    \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i)} \big|$. We vary a
dedicated parameter ($\omega$ in Algorithm 2) in the simulation to force this
ratio to cover a wide range of values. For 179 dataset instances, it ranges
several order of magnitudes with the full distribution of values given in
Table \ref{apd:effect_ratio_distribution}.

\begin{table}[!ht]
    \centering
    \resizebox{\textwidth}{!}{ % Adjusts the table to fit the text width
        \begin{tabular}{|l|r|r|r|r|r|r|r|r|r|r|r|r|}
            \hline
                                   & \textbf{count} & \textbf{mean} & \textbf{std} & \textbf{min} & \textbf{1\%} & \textbf{10\%} & \textbf{25\%} & \textbf{50\%} & \textbf{75\%} & \textbf{90\%} & \textbf{99\%} & \textbf{max} \\ \hline
            \textbf{effect\_ratio} & 179            & 159.6         & 852.4        & 0.004674     & 0.004765     & 0.02483       & 0.1566        & 1.879         & 21.12         & 135.1         & 3394          & 9675         \\ \hline
        \end{tabular}
    }
    \caption{Distribution of the causal effect ratio $\Delta_\mu$ for the experiment on causal effect ratio, results in Figure \ref{fig:effect_ratio_influence}.}\label{apd:effect_ratio_distribution}
\end{table}

\bibliography{biblio}
